---
title: "survAnalysis"
output: html_document
date: "2023-07-12"
---

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("TCGAbiolinks")
```


```{r packages}
require(DESeq2)
require('biomaRt')
require(survival)
require(data.table)
#library(EnhancedVolcano)
library(ggplot2)
library(ggfortify)
library(scales)
library(haven)
library(nnet)
library(simstudy)
library(magrittr)
library(class)
library(dplyr)
library(survminer)
library(TCGAbiolinks)
library(pheatmap)
library(affy)
library(glmnet)
library(tidyr)
library(tibble)
```

This part, I loading the raw htseq-count of transcriptomic data, convert it to read count data and write it to a file.
```{r reading transcriptomic data}
gene.exp <- read.table("../data/TCGA-BLCA.htseq_counts.tsv.gz",
                     header = TRUE,sep = "\t",stringsAsFactors = FALSE)
#colnames(gene.exp) <- gsub("\\.","-",colnames(gene.exp))
gene.exp <- gene.exp[grepl("ENS", gene.exp$Ensembl_ID), ]
gene.exp <- gene.exp[, order(names(gene.exp))]

genecode.id <- gene.exp[,1]
fixed_names <- sapply(strsplit(genecode.id, ".", fixed=T), function(x) x[1])

gene.exp[,1] <- fixed_names

row.names(gene.exp) <- gene.exp[,1]
gene.exp <- gene.exp[-1] #remove double column 
gene.exp <- 2^gene.exp -1
gene.exp <- subset(gene.exp, !(rowSums(gene.exp < 5) == ncol(gene.exp)))

genes.arr <- row.names(gene.exp) #name of genes
write.table(gene.exp, file = "../results/gene_exp_analysis/count_data.csv", sep = "\t", quote = FALSE, row.names = TRUE)
#gene.exp <- as.data.frame(lapply(gene.exp, as.integer))

# checking if the important genes are in the dataset
#"ENSG00000135144" %in% rownames(gene.exp) 
#"ENSG00000143772" %in% rownames(gene.exp) 
#"ENSG00000130332" %in% rownames(gene.exp) 
#"ENSG00000124831" %in% rownames(gene.exp) 
"ENSG00000150764" %in% rownames(gene.exp)
"ENSG00000113657" %in% rownames(gene.exp)
"ENSG00000125895" %in% rownames(gene.exp)
"ENSG00000137693" %in% rownames(gene.exp)
"ENSG00000162520" %in% rownames(gene.exp)

# find the rownumbers of the genes
which(rownames(gene.exp) %in% c("ENSG00000150764", "ENSG00000113657", "ENSG00000125895", 
                                "ENSG00000137693", "ENSG00000162520"))

# make tabke for these genes for heatmap
gene.exp.cut <- gene.exp[c(4202, 5736, 7471, 9276, 10693),]
gene.exp.cut <- as.matrix(gene.exp.cut)
heatmap(gene.exp.cut)
pheatmap(gene.exp.cut)
```

```{r reading clinical data}
clin.data <- read.csv("../data/TCGA-BLCA.survival.tsv",
                      header = TRUE,sep = "\t",stringsAsFactors = FALSE)

clin.data <- clin.data[order(clin.data$sample), ] # order in sample
clin.data$sample <- gsub("\\-",".",clin.data$sample) # replace - with .
color_sample <- ifelse(clin.data$OS == 1, "red", "blue")

table(clin.data$OS) # how many dead (1) and alive (0)
```

```{r reading phenotype data}
phenotype <- read.csv("../data/TCGA-BLCA.GDC_phenotype.tsv.gz",header=TRUE,sep="\t")
phenotype <- phenotype[order(phenotype$submitter_id.samples), ]
phenotype$submitter_id.samples <- gsub("\\-",".",phenotype$submitter_id.samples)
```

Clinical data including stage of patient is here.
```{r Preprocess data}
overlapped.samples <- colnames(gene.exp) #samples from transcriptomic data
clin.data <- clin.data[clin.data$sample %in% overlapped.samples, ]
phenotype <- phenotype[phenotype$submitter_id.samples %in% clin.data$sample, ]
phenotype <- phenotype[!grepl("not reported", phenotype$tumor_stage.diagnoses), ]

gene.exp <- gene.exp[,phenotype$submitter_id.samples]
clin.data <- clin.data[clin.data$sample %in% phenotype$submitter_id.samples, ]
phenotype$tumor_stage.diagnoses <- factor(phenotype$tumor_stage.diagnoses, levels = c("stage i", "stage ii", "stage iii", "stage iv"))
phenotype$tumor_stage.diagnoses <- as.numeric(phenotype$tumor_stage.diagnoses)
clin.data$stage <- phenotype$tumor_stage.diagnoses
phenotype <- phenotype[phenotype$lost_follow_up == "NO",]
clin.data <- clin.data[clin.data$sample %in% phenotype$submitter_id.samples, ]
gene.exp <- gene.exp[,phenotype$submitter_id.samples]
write.table(clin.data, file = "../results/gene_exp_analysis/clinical.csv", sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r}
clin_stage <- read.csv("../results/gene_exp_analysis/clinical.csv",header=TRUE,sep="\t") # ordered
#raw_count <- read.csv("../results/gene_exp_analysis/count_data.csv",header=TRUE,sep = "\t") # the same as gene.exp dataset
```

# Stratifying gene expression data
We removed the stage 1, because all patients survive
```{r}
#separate clinical data by stage
clin_stage_list <- split(clin.data, clin.data$stage)
stage_2 <- clin_stage_list$`2`
stage_3 <- clin_stage_list$`3`
stage_4 <- clin_stage_list$`4`
```

```{r more plots}
#vsdata <- vst(dds_stage2_3, blind=F)
#plotPCA(vsdata, intgroup = "stage")

hist(clin.data$OS.time)
#ggplot(clin.data, aes(x=stage, fill=OS)) + geom_histogram() + facet_wrap(~OS, scales = "free_x")
#ggplot(clin.data, aes(x=OS.time, fill=stage)) + geom_histogram() + facet_wrap(~stage, scales = "free_x")

ggplot(stage_2, aes(x=OS.time, fill=OS)) + geom_histogram() + facet_wrap(~OS, scales = "free_x") +
  ggtitle("Survival of stage 2") +
  xlab("OS time") + ylab("number of samples") + labs(fill = "Survival")

ggplot(stage_3, aes(x=OS.time, fill=OS)) + geom_histogram() + facet_wrap(~OS, scales = "free_x") +
  ggtitle("Survival of stage 3") +
  xlab("OS time") + ylab("number of samples") + labs(fill = "Survival")

ggplot(stage_4, aes(x=OS.time, fill=OS)) + geom_histogram() + facet_wrap(~OS, scales = "free_x") +
  ggtitle("Survival of stage 4") +
  xlab("OS time") + ylab("number of samples") + labs(fill = "Survival")

```

## To do:
- do histogramm of days til death for each stage -> done
- take the gene of intersection and do survival analysis -> done
- but also find important genes/features for each stage
  -> google the genes and look does they have smth to do with the cancer

```{r surv analysis}

#gene.data <- gene.exp |>
#  pivot_longer(cols = everything(), names_to = "sample", values_to = "count")

gene.data <- gene.exp                                           # Duplicate data
gene.data <- tibble::rownames_to_column(gene.data, "row_names") # Apply rownames_to_column
colnames(gene.data)[1] <- "gene_id"                             # rename colname
gene.data  

gene.data["Counts"] <- rowSums(gene.data[,2:293])
gene.data$RowMeans<-as.integer(rowMeans(gene.data[,2:293]))

# making the tables for the significant genes
gene1 <- gene.data[c(4202),]
gene2 <- gene.data[c(5736),]
gene3 <- gene.data[c(7471),]
gene4 <- gene.data[c(9276),]
gene5 <- gene.data[c(10693),]

gene.data1 <- gene1 |>
  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

gene.data2 <- gene2 |>
  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

gene.data3 <- gene3 |>
  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

gene.data4 <- gene4 |>
  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

gene.data5 <- gene5 |>
  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

#gene.Exp <- gene.data |>
#  pivot_longer(cols = starts_with("TCGA"), names_to = "sample", values_to = "count")

# denote which cases have higher / lower expression than median count
#gene.Exp$strata <- ifelse(gene.Exp$count >= gene.Exp$RowMeans, "HIGH", "LOW")

gene.data1$strata <- ifelse(gene.data1$count >= gene.data1$RowMeans, "HIGH", "LOW")
gene.data2$strata <- ifelse(gene.data2$count >= gene.data2$RowMeans, "HIGH", "LOW")
gene.data3$strata <- ifelse(gene.data3$count >= gene.data3$RowMeans, "HIGH", "LOW")
gene.data4$strata <- ifelse(gene.data4$count >= gene.data4$RowMeans, "HIGH", "LOW")
gene.data5$strata <- ifelse(gene.data5$count >= gene.data5$RowMeans, "HIGH", "LOW")

#Assuming your data frame is called "df" and the column you want to remove duplicates from is called "column_name"
#geneExp <- gene.Exp
#unique_values <- unique(geneExp$sample)
# Assuming your data frame is called "df" and the column you want to remove duplicates from is called "column_name"
#geneExp_no_duplicates <- geneExp[!duplicated(geneExp$sample), ]

# Add clinical information to gene.data tables for all stages
gene.Exp1 <- gene.data1
gene.Exp1 <- merge(gene.Exp1, clin_stage, by.x = 'sample', by.y = 'sample')
gene.Exp2 <- gene.data2
gene.Exp2 <- merge(gene.Exp2, clin_stage, by.x = 'sample', by.y = 'sample')
gene.Exp3 <- gene.data3
gene.Exp3 <- merge(gene.Exp3, clin_stage, by.x = 'sample', by.y = 'sample')
gene.Exp4 <- gene.data4
gene.Exp4 <- merge(gene.Exp4, clin_stage, by.x = 'sample', by.y = 'sample')
gene.Exp5 <- gene.data5
gene.Exp5 <- merge(gene.Exp5, clin_stage, by.x = 'sample', by.y = 'sample')

# Add clinical information to gene.data tables for each stage
# stage 2
gene.Exp1.stage2 <- gene.data1
gene.Exp1.stage2 <- merge(gene.Exp1.stage2, stage_2, by.x = 'sample', by.y = 'sample')
gene.Exp2.stage2 <- gene.data2
gene.Exp2.stage2 <- merge(gene.Exp2.stage2, stage_2, by.x = 'sample', by.y = 'sample')
gene.Exp3.stage2 <- gene.data3
gene.Exp3.stage2 <- merge(gene.Exp3.stage2, stage_2, by.x = 'sample', by.y = 'sample')
gene.Exp4.stage2 <- gene.data4
gene.Exp4.stage2 <- merge(gene.Exp4.stage2, stage_2, by.x = 'sample', by.y = 'sample')
gene.Exp5.stage2 <- gene.data5
gene.Exp5.stage2 <- merge(gene.Exp5.stage2, stage_2, by.x = 'sample', by.y = 'sample')

# stage 3
gene.Exp1.stage3 <- gene.data1
gene.Exp1.stage3 <- merge(gene.Exp1.stage3, stage_3, by.x = 'sample', by.y = 'sample')
gene.Exp2.stage3 <- gene.data2
gene.Exp2.stage3 <- merge(gene.Exp2.stage3, stage_3, by.x = 'sample', by.y = 'sample')
gene.Exp3.stage3 <- gene.data3
gene.Exp3.stage3 <- merge(gene.Exp3.stage3, stage_3, by.x = 'sample', by.y = 'sample')
gene.Exp4.stage3 <- gene.data4
gene.Exp4.stage3 <- merge(gene.Exp4.stage3, stage_3, by.x = 'sample', by.y = 'sample')
gene.Exp5.stage3 <- gene.data5
gene.Exp5.stage3 <- merge(gene.Exp5.stage3, stage_3, by.x = 'sample', by.y = 'sample')

# stage 4
gene.Exp1.stage4 <- gene.data1
gene.Exp1.stage4 <- merge(gene.Exp1.stage4, stage_4, by.x = 'sample', by.y = 'sample')
gene.Exp2.stage4 <- gene.data2
gene.Exp2.stage4 <- merge(gene.Exp2.stage4, stage_4, by.x = 'sample', by.y = 'sample')
gene.Exp3.stage4 <- gene.data3
gene.Exp3.stage4 <- merge(gene.Exp3.stage4, stage_4, by.x = 'sample', by.y = 'sample')
gene.Exp4.stage4 <- gene.data4
gene.Exp4.stage4 <- merge(gene.Exp4.stage4, stage_4, by.x = 'sample', by.y = 'sample')
gene.Exp5.stage4 <- gene.data5
gene.Exp5.stage4 <- merge(gene.Exp5.stage4, stage_4, by.x = 'sample', by.y = 'sample')

# survival curves for each gene and each stage -----------

# stage2 ---
sfit1.2 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp1.stage2, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit1.2, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000113657 in stage 2")

sfit2.2 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp2.stage2, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit2.2, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000125895 in stage 2")

sfit3.2 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp3.stage2, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit3.2, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000137693 in stage 2")

sfit4.2 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp4.stage2, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit4.2, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000150764 in stage 2")

sfit5.2 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp5.stage2, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit5.2, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000162520 in stage 2")

# stage3 ---
sfit1.3 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp1.stage3, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit1.3, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000113657 in stage 3")

sfit2.3 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp2.stage3, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit2.3, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000125895 in stage 3")

sfit3.3 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp3.stage3, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit3.3, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000137693 in stage 3")

sfit4.3 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp4.stage3, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit4.3, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000150764 in stage 3")

sfit5.3 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp5.stage3, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit5.3, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000162520 in stage 3")

# stage4 ---
sfit1.4 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp1.stage, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit1.4, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000113657 in stage 4")

sfit2.4 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp2.stage4, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit2.4, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000125895 in stage 4")

sfit3.4 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp3.stage4, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit3.4, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000137693 in stage 4")

sfit4.4 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp4.stage4, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit4.4, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000150764 in stage 4")

sfit5.4 <- survfit(Surv(OS.time, OS)~ strata, data=gene.Exp5.stage4, type = "kaplan-meier", 
            conf.type = "log")
autoplot(sfit5.4, conf.int = F, main="Kaplan-Meier-estimate for the gene ENSG00000162520 in stage 4")
```
